---
title: "Employee Attrition Exploration & Analyses"
subtitle: "A People Analytics Practice Project"
author: "Keefer Yip"
format: 
  html:
    theme: sandstone
    toc: true
execute:
    warning: false
    message: false
editor: visual
---

## Introduction

The following document utilizes data from the [IBM HR Analytics Employee Attrition Dataset](https://www.kaggle.com/datasets/pavansubhasht/ibm-hr-analytics-attrition-dataset). The primary purpose was to highlight and summarize skills in R, primarily things such as exploratory data analysis, as well as data cleaning, manipulation, and visualization.

This project also answers some brief questions regarding attrition and other metrics, while posing new areas for discussion.

It is important to note that units are not specified in the dataset for variables such as distance and income. When applicable, factors were converted using explicit scales from the information provided.

## Setup

```{r}
#| label: setup

# Load necessary packages
library(tidyverse)
library(ggplot2)
library(janitor)
```

### Importing Data

```{r}
#| label: data-import

# Load the dataset
hr_data <- read_csv("data/WA_Fn-UseC_-HR-Employee-Attrition.csv")
```

```{r}
#| include: false
# Examining an overview of the dataset
glimpse(hr_data)
```

## Cleaning

```{r}
#| label: data cleaning
# Standardized column names
hr_data <- janitor::clean_names(hr_data) %>% 

# Removed irrelevant variables 
  select(-any_of(c("employee_count", "standard_hours", "over18")))

# Created first data branch, for feature analysis
hr_features <- hr_data %>% 
  select(employee_number, attrition, age, 
         years_at_company, over_time, distance_from_home, 
         ends_with("_satisfaction"), work_life_balance) 


# Creating vector for following ordinal factor creation, all have the same factor levels and labels
satisfaction_vars <- c("job_involvement", "job_satisfaction", 
                       "environment_satisfaction", "relationship_satisfaction")

# Created second data branch for EDA
hr_eda <- hr_data %>% 
  mutate(
    # Ordinal factor creation for satisfaction variables
    across(
      all_of(satisfaction_vars),
      ~ factor(.,
               levels = 1:4,
               labels = c("Low", "Medium", "High", "Very High"),
               ordered = TRUE)
    ),
    # Ordinal factor for work life balance (has different labels than satisfaction)
    work_life_balance = factor(
      work_life_balance,
      levels = 1:4,
      labels = c("Bad", "Good", "Better", "Best"),
      ordered = TRUE
    ),
    # Ordinal factor for education 
    education = factor(
      education,
      levels = 1:5,
      labels = c("Below College", "College",
                 "Bachelor", "Master", "Doctor"),
      ordered = TRUE
    ),
    # Nominal categorical variables
    across(
      c(attrition, business_travel, department, education_field,
        gender, job_role, marital_status, over_time),
      as.factor
    )
  )
```

## Analyses

For the analyses, I wanted to explore different avenues that may help us get a better understanding of where attrition was more likely to occur, in turn allowing us to highlight groups that were at a bigger risk.

```{r}
#| label: Preparation for analyses

# Created a new table highlighting employees who attritted, selecting some specific variables
attrition_data <- hr_eda %>% 
  filter(attrition == "Yes") %>% 
  select(employee_number, distance_from_home, education, environment_satisfaction, 
         job_involvement, job_level, job_satisfaction, 
         num_companies_worked:last_col())

```

### Engagement Insights

Upon my initial glance at the data, the first question that caught my eye was about the relationship between job involvement and satisfaction, which I called "job engagement".

\[Add more about why I chose these variables?\]

```{r}
#| label: Job Involvement and Satisfaction graph

# Created a new table to display the attrition rates of each job involvement and satisfaction pair
engagement_df <- hr_eda %>% 
  group_by(job_involvement, job_satisfaction) %>% 
  summarize(attrition_rate = round((mean(attrition == "Yes")) * 100, 2))

# Created theme for following plots
graph_theme <- theme_classic() +
  theme(
    axis.title = element_text(size = 12),
    legend.text = element_text(size = 12))

# Created heatmap showcasing the relationship between the two variables
engagement_heatmap <- ggplot(engagement_df) + 
  geom_tile(aes(x = job_involvement, 
                y = job_satisfaction, 
                fill = attrition_rate)) +
  scale_fill_gradient() +
  labs(x = "Job Involvement", 
       y = "Job Satisfaction",
       title = "Job Engagement Heatmap",
       fill = "Attrition Rate") +
  theme_classic() +
  graph_theme

engagement_heatmap
```

The heatmap shows that attrition rates are lowest at the extreme highs of job involvement and satisfaction. In fact, the rate for employees with both very high job involvement and satisfaction is `r min(engagement_df$attrition_rate)`%. As shown in the graphic, as these engagement levels decrease, attrition rates tend to increase.

### Commute and Satisfaction

The following graphs outline the relationship between several factors: commute distance, satisfaction, and attrition.

I decided to take a look into insights of potential areas that may serve as explanation to why employees engage in attrition.

To do so, I created a new feature, **satisfaction_index**, which took the average value (from a scale of 1 \[Low\] to 4 \[Very High\]) from all other satisfaction variables, including environment, job, and relationship satisfaction, as well as work life balance.

\*Note that commute distance does not have a unit because the data set is not specific.

```{r}
#| label: Organizing Data

# Engineered new feature, satisfaction_index that takes the average of the satisfaction variables
hr_features <- hr_features %>% 
  mutate(
    satisfaction_index = (
      environment_satisfaction + 
        job_satisfaction +  
        relationship_satisfaction + 
        work_life_balance
      ) / 4
    )

# Graph the relationship between satisfaction and attrition
ggplot(hr_features, aes(satisfaction_index, attrition)) +
  geom_boxplot() +
  labs(
    title = "Satisfaction Index and Attrition", 
    x = "Satisfaction Index Score", 
    y = "Attrition") +
  graph_theme

# Graph the relationship between commute distance and attrition
ggplot(hr_features, aes(distance_from_home, attrition)) +
  geom_boxplot() +
  labs(
    title = "Commute and Attrition", 
    x = "Commute Distance", 
    y = "Attrition") +
  graph_theme

# Graph the relationship between commute distance and satisfaction
ggplot(hr_features, aes(distance_from_home, satisfaction_index)) +
  geom_jitter(alpha = 0.2, width = 0.3) +
  geom_smooth(method = "loess") +
  labs(
    title = "Commute and Satisfaction", 
    x = "Commute Distance", 
    y = "Satisfaction Index Score") +
  graph_theme
```

The graphs mentioned above show that satisfaction is lower and commute distance slightly higher for employees experiencing attrition, although there isn't a strong relationship between satisfaction and commute distance themselves.

### Education Insights

```{r}
# Created a new table to gain insight on employees based on their education level
education_data <- attrition_data %>% 
  select(employee_number, education:job_satisfaction, percent_salary_hike, 
         total_working_years, years_at_company:years_since_last_promotion)

# Finding which level of education has had the highest attrition rate
ed_df <- hr_eda %>% 
group_by(education) %>% 
  summarize(attrition_rate = round((mean(attrition == "Yes")) * 100, 2))

# Displaying the attrition rates of each education level
knitr::kable(ed_df)
```

As seen, the education level with the highest rate of attrition was the employees who had a "Below College" level, with a rate of `r max(ed_df$attrition_rate)`%. Conversely, the education level with the lowest rate of attrition was the employees with a "Doctor" level education, with a rate of `r min(ed_df$attrition_rate)`%.

```{r}
#| include: false
range(hr_eda$years_since_last_promotion)

```

#### Graph of Breakdown by Years Since Last Promotion

```{r}
#| label: Education x Promotion Graphing

# Created graph showing employees who attritted based on years since their last promotion
promo_graph_count <- education_data %>% 
  mutate(years_bucket = cut(years_since_last_promotion,
                            breaks = c(0, 3, 6, 9, 12, 15),
                            labels = c("0-3", "4-6", "7-9", "10-12", "13-15"),
                            include.lowest = TRUE)) %>%
  ggplot(aes(years_bucket, 
             fill = education)) + 
  scale_x_discrete("Years Since Last Promotion") +
  scale_y_continuous("Count", expand = c(0,0)) +
  geom_bar(position = "dodge", width = .8) +
  labs(title = "Breakdown of Attrition by Years Since Last Promotion", 
       fill = "Education Level") +
  graph_theme +
  scale_fill_brewer(palette = "Blues")
      
promo_graph_count
```

#### Graph displaying count above, graph displaying proportion below

```{r}
#| label: Education proportion graph
#| code-fold: true

# Created table and graph to show the proportion of employees who attritted based on years since their last promotion
promo_prop_df <- education_data %>% 
  mutate(years_bucket = cut(years_since_last_promotion,
                            breaks = c(0, 3, 6, 9, 12, 15),
                            labels = c("0-3", "4-6", "7-9", "10-12", "13-15"),
                            include.lowest = TRUE)) %>%
  count(years_bucket, education) 

# Table to display the breakdown into each bucket
knitr::kable(promo_prop_df)

promo_graph_prop <- promo_prop_df %>% 
  group_by(years_bucket) %>% 
  mutate(prop = n / sum(n)) %>% 
  ggplot(aes(years_bucket,
             prop,
             fill = education)) + 
  scale_x_discrete("Years Since Last Promotion") +
  scale_y_continuous("Proportion", expand = c(0,0)) +
  geom_col(position = "dodge", width = .8) +
  labs(title = "Breakdown of Attrition by Years Since Last Promotion", 
       fill = "Education Level") +
  graph_theme + 
  scale_fill_brewer(palette = "Blues")
      
promo_graph_prop
```

### Early Career Burnout?

Are younger/newer employees with overtime and lower satisfaction more likely to leave?

Multi-layered hypothesis:

Early-career employees leave more.

Overtime increases attrition.

Low satisfaction and overtime lead to extreme attrition.

```{r}
#| label: EDA before analyses and graphs

# Created segments for employee career stages and satisfaction 
hr_features <- hr_features %>% 
  mutate(
    career_stage = case_when(
      years_at_company <= 2 ~ "Early Career",
      years_at_company <= 7 ~ "Mid Career",
      TRUE ~ "Established"),
    satisfaction_group = case_when(
      satisfaction_index <= 2 ~ "Low",
      satisfaction_index <= 3 ~ "Medium",
      TRUE ~ "High"
    ),
    # Turn segments into ordered factors
    career_stage = factor(
      career_stage, 
      levels = c("Early Career", "Mid Career", "Established")
    ),
    satisfaction_group = factor(
      satisfaction_group, 
      levels = c("Low", "Medium", "High")
    )
  )

# Examining the attrition rate across career stages
hr_features %>% 
  group_by(career_stage) %>% 
  summarize(
    n = n(),
    attrition_rate = round((mean(attrition == "Yes")) * 100, 2) 
  )%>% 
  arrange(desc(attrition_rate)) %>% 
knitr::kable()

# Examining attrition rate across career stages and OT
hr_features %>% 
  group_by(career_stage, over_time) %>% 
  summarize(
    n = n(),
    attrition_rate = round((mean(attrition == "Yes")) * 100, 2)
  ) %>% 
  arrange(desc(attrition_rate)) %>% 
knitr::kable()
```

We see that employees in the early stages of their career and have overtime are more likely to engage in attrition. However, I wanted to take it a step further by examining the satisfaction of these groups.

```{r}
#| label: OT Analyses

# Created a risk table to showcase the 5 groups with the highest attrition rates
risk_table <- hr_features %>% 
  group_by(career_stage, over_time, satisfaction_group) %>% 
  summarize(
    n = n(), 
    attrition_rate = round((mean(attrition == "Yes")) * 100, 2)
    ) %>% 
  ungroup() %>% 
  slice_max(attrition_rate, n = 5)

knitr::kable(risk_table)

# Create a graph to display effects of overtime on different career stages
overtime_graph <- ggplot(hr_features, aes(x = career_stage, fill = attrition)) +
  geom_bar(position = "fill") +
  facet_grid(~ over_time, 
             labeller = labeller(
               over_time = c("No" = "No Overtime", 
                             "Yes" = "Overtime")
               )
             ) + 
  labs(
    title = "Attrition Patterns Across Career Stages and Overtime",
    x = "Career Stage",
    y = "Proportion",
    fill = "Attrition"
    ) +
  graph_theme +
  scale_fill_manual(values = c("No" = "grey80", "Yes" = "#D55E00"))

overtime_graph

# Deeper dive into early career employees and their satisfaction
early_career <- hr_features %>% 
  filter(career_stage == "Early Career") 

early_overtime_graph <- ggplot(early_career, 
                               aes(x = satisfaction_group, fill = attrition)) +
  geom_bar(position = "fill") +
  facet_wrap(~ over_time, 
             labeller = labeller(
               over_time = c("No" = "No Overtime", 
                             "Yes" = "Overtime")
               )
             ) +
  labs(
    title = "Early Career Attrition: Effects of Satisfaction and Overtime",
    x = "Satisfaction",
    y = "Proportion",
    fill = "Attrition"
    ) +
  graph_theme +
  scale_fill_manual(values = c("No" = "grey80", "Yes" = "#D55E00"))

early_overtime_graph
```

## Concluding Thoughts
