---
title: "Employee Attrition Exploration & Analyses"
subtitle: "A People Analytics Practice Project"
author: "Keefer Yip"
format: 
  html:
    theme: sandstone
    toc: true
execute:
    warning: false
    message: false
editor: visual
---

## Setup

```{r}
#| label: setup

# Load necessary packages
library(tidyverse)
library(ggplot2)
library(janitor)
```

### Importing Data

```{r}
#| label: data-import

# Load the dataset
hr_data <- read_csv("data/WA_Fn-UseC_-HR-Employee-Attrition.csv")
```

### Initial Inspection

```{r}
# Examining an overview of the dataset
glimpse(hr_data)
```

# Cleaning

```{r}
#| label: data cleaning
# Standardized column names
hr_data <- janitor::clean_names(hr_data) %>% 

# Removed standardized columns and irrelevant variables to following analyses
  select(-any_of(c("employee_count", "standard_hours", "over18")))

# Created first data branch, for feature analysis
hr_features <- hr_data %>% 
  select(employee_number, attrition, age, 
         years_at_company, over_time, distance_from_home, 
         ends_with("_satisfaction"), work_life_balance) 


# Creating vector for following ordinal factor creation, all have the same factor levels and labels
satisfaction_vars <- c("job_involvement", "job_satisfaction", 
                       "environment_satisfaction", "relationship_satisfaction")

# Created second data branch for EDA
hr_eda <- hr_data %>% 
  mutate(
    # Ordinal factor creation for satisfaction variables
    across(
      all_of(satisfaction_vars),
      ~ factor(.,
               levels = 1:4,
               labels = c("Low", "Medium", "High", "Very High"),
               ordered = TRUE)
    ),
    # Ordinal factor for work life balance (has different labels than satisfaction)
    work_life_balance = factor(
      work_life_balance,
      levels = 1:4,
      labels = c("Bad", "Good", "Better", "Best"),
      ordered = TRUE
    ),
    # Ordinal factor for education 
    education = factor(
      education,
      levels = 1:5,
      labels = c("Below College", "College",
                 "Bachelor", "Master", "Doctor"),
      ordered = TRUE
    ),
    
    # Nominal categorical variables
    across(
      c(attrition, business_travel, department, education_field,
        gender, job_role, marital_status, over_time),
      as.factor
    )
  )
```

# Analyses

```{r}
#| label: analyses

# Created a new table highlighting employees who attritted, selecting some specific variables
attrition_data <- hr_eda %>% 
  filter(attrition == "Yes") %>% 
  select(employee_number, distance_from_home, education, environment_satisfaction, 
         job_involvement, job_level, job_satisfaction, 
         num_companies_worked:last_col())

```

## Engagement Insights

```{r}
#| label: Job Involvement and Satisfaction graph

# Created a new table to display the attrition rates of each job involvement and satisfaction pair
engagement_df <- hr_eda %>% 
  group_by(job_involvement, job_satisfaction) %>% 
  summarize(attrition_rate = round((mean(attrition == "Yes")) * 100, 2))

# Created theme for following plots
graph_theme <- theme_classic() +
  theme(
    axis.title = element_text(size = 12),
    legend.text = element_text(size = 12))

# Created heatmap showcasing the relationship between the two variables
engagement_heatmap <- ggplot(engagement_df) + 
  geom_tile(aes(x = job_involvement, 
                y = job_satisfaction, 
                fill = attrition_rate)) +
  scale_fill_gradient() +
  labs(x = "Job Involvement", 
       y = "Job Satisfaction",
       title = "Job Engagement Heatmap",
       fill = "Attrition Rate") +
  theme_classic() +
  graph_theme

engagement_heatmap
```

## Commute and Satisfaction

How does Commute influence Satisfaction?

I took a snapshot into various insights of potential areas that may serve as some explanation to why employees engage in attrition. The following graphs outline the relationship between several factors: commute distance, satisfaction, and attrition. \[REWORD\]

Created new feature, satisfaction_index... (Important to discuss because I use it later as well)

```{r}
#| label: Organizing Data

# Engineered new feature, satisfaction_index that takes the average of the satisafction variables
hr_features <- hr_features %>% 
  mutate(
    satisfaction_index = (
      environment_satisfaction + 
        job_satisfaction +  
        relationship_satisfaction + 
        work_life_balance
      ) / 4
    )

# Graph the relationship between satisfaction and attrition
ggplot(hr_features, aes(satisfaction_index, attrition)) +
  geom_boxplot()

# Graph the relationship between commute distance and attrition
ggplot(hr_features, aes(distance_from_home, attrition)) +
  geom_boxplot()


# Graph the relationship between commute distance and satisfaction
ggplot(hr_features, aes(distance_from_home, satisfaction_index)) +
  geom_jitter(alpha = 0.2, width = 0.3) +
  geom_smooth(method = "loess")
```

The graphs mentioned above show that satisfaction is lower and commute distance slightly higher for employees experiencing attrition, although there isn't a strong relationship between satisfaction and commute distance themselves.

## Education Insights

```{r}
# Created a new table to gain insight on employees based on their education level
education_data <- attrition_data %>% 
  select(employee_number, education:job_satisfaction, percent_salary_hike, 
         total_working_years, years_at_company:years_since_last_promotion)

# Finding which level of education has had the highest attrition rate
ed_df <- hr_eda %>% 
group_by(education) %>% 
  summarize(attrition_rate = round((mean(attrition == "Yes")) * 100, 2))

# Displaying the attrition rates of each education level
knitr::kable(ed_df)
```

As seen, the education level with the highest rate of attrition was the employees who had a "Below College" level, with a rate of `r max(ed_df$attrition_rate)`%. Conversely, the education level with the lowest rate of attrition was the employees with a "Doctor" level education, with a rate of `r min(ed_df$attrition_rate)`%.

```{r}
#| include: false
range(hr_eda$years_since_last_promotion)

```

#### Graph of Breakdown by Years Since Last Promotion

```{r}
#| label: Graphing

# Created graph to show the count of employees who attritted based on years since their last promotion
promo_graph_count <- education_data %>% 
  mutate(years_bucket = cut(years_since_last_promotion,
                            breaks = c(0, 3, 6, 9, 12, 15),
                            labels = c("0-3", "4-6", "7-9", "10-12", "13-15"),
                            include.lowest = TRUE)) %>%
  ggplot(aes(years_bucket, 
             fill = education)) + 
  scale_x_discrete("Years Since Last Promotion") +
  scale_y_continuous("Count", expand = c(0,0)) +
  geom_bar(position = "dodge", width = .8) +
  labs(title = "Breakdown of Attrition by Years Since Last Promotion", 
       fill = "Education Level") +
  graph_theme +
  scale_fill_brewer(palette = "Accent")
      
promo_graph_count
```

#### Graph displaying count above, graph displaying proportion below

```{r}
#| code-fold: true

# Created table and graph to show the proportion of employees who attritted based on years since their last promotion
promo_prop_df <- education_data %>% 
  mutate(years_bucket = cut(years_since_last_promotion,
                            breaks = c(0, 3, 6, 9, 12, 15),
                            labels = c("0-3", "4-6", "7-9", "10-12", "13-15"),
                            include.lowest = TRUE)) %>%
  count(years_bucket, education) 

# Table to display the breakdown into each bucket
knitr::kable(promo_prop_df)

promo_graph_prop <- promo_prop_df %>% 
  group_by(years_bucket) %>% 
  mutate(prop = n / sum(n)) %>% 
  ggplot(aes(years_bucket,
             prop,
             fill = education)) + 
  scale_x_discrete("Years Since Last Promotion") +
  scale_y_continuous("Proportion", expand = c(0,0)) +
  geom_col(position = "dodge", width = .8) +
  labs(title = "Breakdown of Attrition by Years Since Last Promotion", 
       fill = "Education Level") +
  graph_theme + 
  scale_fill_brewer(palette = "Accent")
      
promo_graph_prop
```

## Early Career Burnout?

Are younger/newer employees with overtime and lower satisfaction more likely to leave?

Multi-layered hypothesis:

Early-career employees leave more.

Overtime increases attrition.

Low satisfaction and overtime lead to extreme attrition.

```{r}
#| label: EDA before analyses and graphs

# Created segments for employee career stages and satisfaction 
hr_features <- hr_features %>% 
  mutate(
    career_stage = case_when(
      years_at_company <= 2 ~ "Early Career",
      years_at_company <= 7 ~ "Mid Career",
      TRUE ~ "Established"),
      satisfaction_group = case_when(
        satisfaction_index <= 2 ~ "Low",
        satisfaction_index <= 3 ~ "Medium",
        TRUE ~ "High"
      )
    )

# Examining the attrition rate across career stages
hr_features %>% 
group_by(career_stage) %>% 
  summarize(n = n(),
            attrition_rate = mean(attrition == "Yes")) %>% 
  arrange(desc(attrition_rate))

# Examining attrition rate across career stages and OT
hr_features %>% 
  group_by(career_stage, over_time) %>% 
  summarize(
    n = n(),
    attrition_rate = mean(attrition == "Yes")
  ) %>% 
  arrange(desc(attrition_rate))


```

```{r}
#| label: Analyses


```
