---
title: "Employee Attrition Exploration & Analyses"
subtitle: "A People Analytics Practice Project"
author: "Keefer Yip"
format: 
  html:
    theme: sandstone
    toc: true
execute:
    warning: false
    message: false
editor: visual
---

## Setup

```{r}
#| label: setup

# Load necessary packages
library(tidyverse)
library(ggplot2)
library(janitor)
```

### Importing Data

```{r}
#| label: data-import

# Load the dataset
hr_data <- read_csv("data/WA_Fn-UseC_-HR-Employee-Attrition.csv")
```

### Initial Inspection

```{r}
# Examining an overview of the dataset
glimpse(hr_data)
```

# Cleaning

```{r}
#| label: data cleaning
# Standardized column names
hr_data <- janitor::clean_names(hr_data) %>% 

# Removed standardized columns and irrelevant variables to following analyses
  select(-any_of(c("employee_count", "standard_hours", "over18")))


hr_data <- hr_data %>% 
# Convert appropriate variables into ordinal factors  
  mutate(job_involvement = factor(job_involvement, 
                                  levels = 1:4, 
                                  labels = c("Low", "Medium", "High", "Very High")), 
         job_satisfaction = factor(job_satisfaction, 
                                   levels = 1:4, 
                                   labels = c("Low", "Medium", "High", "Very High")),
         education = factor(education, 
                                  levels = 1:5,
                                  labels = c("Below College", "College", 
                                             "Bachelor", "Master", "Doctor"))
  ) %>% 
# Convert appropriate variables into nominal factors
  mutate(across(c(attrition, business_travel, department, education_field, 
                  gender, job_role, marital_status, over_time), as.factor))
```

# Analyses

```{r}
#| label: analyses

# Created a new table highlighting employees who attritted, selecting some specific variables
attrition_data <- hr_data %>% 
  filter(attrition == "Yes") %>% 
  select(employee_number, distance_from_home, education, environment_satisfaction, 
         job_involvement, job_level, job_satisfaction, 
         num_companies_worked:last_col())

```

## Engagement Insights

```{r}
#| label: Job Involvement and Satisfaction graph

# Created a new table to display the attrition rates of each job involvement and satisfaction pair
engagement_df <- hr_data %>% 
  group_by(job_involvement, job_satisfaction) %>% 
  summarize(attrition_rate = round((mean(attrition == "Yes")) * 100, 2))

# Created theme for following plots
graph_theme <- theme_classic() +
  theme(
    axis.title = element_text(size = 12),
    legend.text = element_text(size = 12))

# Created heatmap showcasing the relationship between the two variables
engagement_heatmap <- ggplot(engagement_df) + 
  geom_tile(aes(x = job_involvement, 
                y = job_satisfaction, 
                fill = attrition_rate)) +
  scale_fill_gradient() +
  labs(x = "Job Involvement", 
       y = "Job Satisfaction",
       title = "Job Engagement Heatmap",
       fill = "Attrition Rate") +
  theme_classic() +
  graph_theme

engagement_heatmap
```

## Education Insights

```{r}
# Created a new table to gain insight on employees based on their education level
education_data <- attrition_data %>% 
  select(employee_number, education:job_satisfaction, percent_salary_hike, 
         total_working_years, years_at_company:years_since_last_promotion)

# Finding which level of education has had the highest attrition rate
ed_df <- hr_data %>% 
group_by(education) %>% 
  summarize(attrition_rate = round((mean(attrition == "Yes")) * 100, 2))

# Displaying the attrition rates of each education level
knitr::kable(ed_df)
```

As seen, the education level with the highest rate of attrition was the employees who had a "Below College" level, with a rate of `r max(ed_df$attrition_rate)`%. Conversely, the education level with the lowest rate of attrition was the employees with a "Doctor" level education, with a rate of `r min(ed_df$attrition_rate)`%.

#### Graph of Breakdown by Years Since Last Promotion

```{r}
#| label: Graphing

# Created graph to show the count of employees who attritted based on years since their last promotion
promo_graph_count <- education_data %>% 
  mutate(years_bucket = cut(years_since_last_promotion,
                            breaks = c(0, 3, 6, 9, 12, 15),
                            labels = c("0-3", "4-6", "7-9", "10-12", "13-15"),
                            include.lowest = TRUE)) %>%
  ggplot(aes(years_bucket, 
             fill = education)) + 
  scale_x_discrete("Years Since Last Promotion") +
  scale_y_continuous("Count", expand = c(0,0)) +
  geom_bar(position = "dodge", width = .8) +
  labs(title = "Breakdown of Attrition by Years Since Last Promotion", 
       fill = "Education Level") +
  graph_theme +
  scale_fill_brewer(palette = "Accent")
      
promo_graph_count
```

#### Graph displaying count above, graph displaying proportion below

```{r}
#| code-fold: true

# Created table and graph to show the proportion of employees who attritted based on years since their last promotion
promo_prop_df <- education_data %>% 
  mutate(years_bucket = cut(years_since_last_promotion,
                            breaks = c(0, 3, 6, 9, 12, 15),
                            labels = c("0-3", "4-6", "7-9", "10-12", "13-15"),
                            include.lowest = TRUE)) %>%
  count(years_bucket, education) 

# Table to display the breakdown into each bucket
knitr::kable(promo_prop_df)

promo_graph_prop <- promo_prop_df %>% 
  mutate(prop = n / sum(n)) %>% 
  ggplot(aes(years_bucket, 
             prop,
             fill = education)) + 
  scale_x_discrete("Years Since Last Promotion") +
  scale_y_continuous("Proportion", expand = c(0,0)) +
  geom_col(position = "dodge", width = .8) +
  labs(title = "Breakdown of Attrition by Years Since Last Promotion", 
       fill = "Education Level") +
  graph_theme + 
  scale_fill_brewer(palette = "Accent")
      
promo_graph_prop
```

```{r}

```
